Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-17 ms.tex
Added note about arbitrary ngal error in Zehavi+11, Added note about Zehavi+11 using FoF halos whereas we use virial.

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced decorated Halo Occupation Distribution (HOD) 
  model using SDSS DR7 measurements of projected galaxy clustering, $\wprp (r_{\rm p})$ 
  of galaxies in r-band luminosity-threshold samples. 
  The decorated HOD is a model for the galaxy--halo
  connection that augments the traditional HOD by allowing for the possibility of 
  {\em assembly bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides mass, $\mvir$. We demonstrate that it
  is not possible to rule out galaxy assembly bias using DR7 measurements of galaxy 
  clustering alone. Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias. Our analyses favor models 
  in which high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite galaxy assembly bias is significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. In addition to presenting the 
  constraints on HOD models that accommodate assembly bias, our analysis 
  includes numerous improvements over previous analyses of this data set and 
  supersedes existing constraints even on standard HOD parameters.
\end{abstract}

@@

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the difference
between the projected two-point function predicted by the mock catalog and the measured value in the i$^{\rm th}$
separation bin (of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix of the measurements,
and the last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference between the predicted and measured
galaxy number density. The error on the galaxy number density assumes Poisson statistics for both the measured and predicted galaxy number densities.

@@

The most important detail of this analysis is the priors on the parameters. In all analyses
discussed in this paper, we adopt priors that are uniform distributions over the intervals
specified in Table~\ref{table:priors}. In the case of the assembly bias parameters
$A_{\mathrm{cen}}$ and $A_{\mathrm{sat}}$, the priors have strict
boundaries. Mathematically, these parameters must satisfy $-1 \le A_{\mathrm{cen,sat}} \le 1$.
Physical considerations require parameter $\sigma_{\log M} > 0$. All other
priors have a negligible influence on the posterior aside from $\log M_0$.
We find that $\log M_0$ is often very poorly constrained by clustering data and
priors on $\log M_0$ can have a non-negligible influence on inferred parameters.

@@

Our parameter constraints can be compared to the standard HOD analysis performed by \citet{zehavi_etal11} 
by examining the top two rows in each luminosity threshold grouping in Table~\ref{table:parameters}. 
The inferred parameters from our standard analyses differ in several ways from the \citet{zehavi_etal11}
analysis. Firstly, in our re-analysis of the projected clustering data, we generally find all mass scales to
be slightly higher than in the work of \citet{zehavi_etal11}. This difference is primarily due to the slightly
different cosmological model used in this work as compared to that of \citet{zehavi_etal11}.
The most important differences are in the values of $\Omega_{\rm M}$, and $\sigma_8$.
\citet{zehavi_etal11} assumed $\Omega_{\rm M}=0.25$ and $\sigma_8=0.8$,
whereas in the present work, we use the BolshoiP simulation in which $\Omega_{\rm M}=0.307$ and $\sigma_8=0.82$.
Slightly larger mass scales are necessary in an analysis with higher $\Omega_{\rm M}$ and $\sigma_8$ in order
to maintain galaxy number densities fixed with larger halo number densities.

@@

A second noteworthy difference between the present work and that of
\citet{zehavi_etal11} is that we find many parameters to be notably
more poorly constrained. At the lower luminosity thresholds, for
example, we constrain $\log (M_{\rm min})$ and $\sigma_{\log M}$ with
several times lower precision than \citet{zehavi_etal11}. We do not
show our constraints on $\log (M_0)$ as they are very poor, with
1-sigma constraints $\gta 1$~dex for all samples. In several cases,
the constraint on $\log (M_0)$ is determined by the prior given in
Table~\ref{table:priors}. This is in stark contrast to several of the
results of \citet{zehavi_etal11}. For example, for the threshold
sample with $M_r < -19.5$ ($M_r < -20.5$), \citet{zehavi_etal11} quote
$\log (M_0) = 12.23 \pm 0.17$ ($12.35 \pm 0.24$), whereas we infer
$\log (M_0) = 11.38^{+0.95}_{-1.57}$
($11.19^{+0.89}_{-1.39}$). Examining the form of Eq.~(\ref{eq:nsat}),
it is sensible that the parameter $\log (M_0)$ should be unconstrained
at the lower end, because the value of $M_0$ does not alter the
predicted satellite number once $M_0 \ll M_1$. Therefore, it seems
likely that the tighter constraints quoted by \citet{zehavi_etal11}
must be an error.

>>>>

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced decorated Halo Occupation Distribution (HOD) 
  model using SDSS DR7 measurements of projected galaxy clustering, $\wprp (r_{\rm p})$ 
  of galaxies in r-band luminosity-threshold samples. 
  The decorated HOD is a model for the galaxy--halo
  connection that augments the traditional HOD by allowing for the possibility of 
  {\em assembly bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides mass, $\mvir$. We demonstrate that it
  is not possible to rule out galaxy assembly bias using DR7 measurements of galaxy 
  clustering alone. Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias. These samples favor models 
  in which high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite galaxy assembly bias is significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. In addition to presenting the first 
  constraints on HOD models that accommodate assembly bias, our analysis 
  includes numerous improvements over previous analyses of this data set and 
  supersedes previously published results, even in the case of a standard HOD analysis.
\end{abstract}

@@

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the difference
between the projected two-point function predicted by the mock catalog and the measured value in the i$^{\rm th}$
separation bin (of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix of the measurements,
and the last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference between the predicted and measured
galaxy number density. In Eq.~(\ref{eq:chisquare}), the we adopt the Einstein summation convention and sum over the 
dummy indices $i$ and $j$. The error on the galaxy number density assumes Poisson statistics for both the measured and predicted galaxy number densities.

@@

The most important detail of this analysis is the priors on the parameters. In all analyses
discussed in this paper, we adopt priors that are uniform distributions over the intervals
specified in Table~\ref{table:priors}. In the case of the assembly bias parameters
$A_{\mathrm{cen}}$ and $A_{\mathrm{sat}}$, the priors have strict
boundaries. Mathematically, these parameters must satisfy $-1 \le A_{\mathrm{cen,sat}} \le 1$.
Physical considerations require parameter $\sigma_{\log M} > 0$. All other
priors have a negligible influence on the posterior aside from $\log M_0$ and 
$\sigma_{\log M}$. We find that $\log M_0$ is often very poorly constrained by clustering data and
priors on $\log M_0$ can have a non-negligible influence on inferred parameters. The parameter 
$\sigma_{\log M}$ is poorly constrained for the $\magr < -19$ sample.

@@

Our parameter constraints can be compared to the standard HOD analysis performed by \citet{zehavi_etal11} 
by examining the top two rows in each luminosity threshold grouping in Table~\ref{table:parameters}. In all 
cases, we quote the medians of our posteriors as our central values and our, 
generally asymmetric, errorbars, give the 16$^{\rm th}$ and 
84$^{\rm th}$ percentiles of the posterior samples so as to correspond roughly with ``$1-\sigma$" errors. 
The inferred parameters from our standard analyses differ in several ways from the \citet{zehavi_etal11}
analysis. Firstly, in our re-analysis of the projected clustering data, we generally find all mass scales to
be slightly higher than in the work of \citet{zehavi_etal11}. This difference is primarily due to the slightly
different cosmological model used in this work as compared to that of \citet{zehavi_etal11}.
The most important differences are in the values of $\Omega_{\rm M}$, and $\sigma_8$.
\citet{zehavi_etal11} assumed $\Omega_{\rm M}=0.25$ and $\sigma_8=0.8$,
whereas in the present work, we use the BolshoiP simulation in which $\Omega_{\rm M}=0.307$ and $\sigma_8=0.82$.
Slightly larger mass scales are necessary in an analysis with higher $\Omega_{\rm M}$ and $\sigma_8$ in order
to maintain galaxy number densities fixed with larger halo number densities. A detailed comparison between our central 
values and those of \citet{zehavi_etal11} is further confounded by the fact that we use a virial halo definition, 
whereas the analytic model of \citet{zehavi_etal11} 
has been calibrated to friend-of-friends halo masses (Z. Zheng, Private Communication).

@@

A second noteworthy difference between the present work and that of
\citet{zehavi_etal11} is that we find many parameters to be notably
more poorly constrained. At the lower luminosity thresholds, for
example, we constrain $\log (M_{\rm min})$ and $\sigma_{\log M}$ with
several times lower precision than \citet{zehavi_etal11}. We do not
show our constraints on $\log (M_0)$ as they are very poor, with
1-sigma constraints $\gta 1$~dex for all samples. In several cases,
the constraint on $\log (M_0)$ is determined by the prior given in
Table~\ref{table:priors}. This is in stark contrast to several of the
results of \citet{zehavi_etal11}. For example, for the threshold
sample with $M_r < -19.5$ ($M_r < -20.5$), \citet{zehavi_etal11} quote
$\log (M_0) = 12.23 \pm 0.17$ ($12.35 \pm 0.24$), whereas we infer
$\log (M_0) = 11.38^{+0.95}_{-1.57}$
($11.19^{+0.89}_{-1.39}$). Examining the form of Eq.~(\ref{eq:nsat}),
it is sensible that the parameter $\log (M_0)$ should be unconstrained
at the lower end, because the value of $M_0$ does not alter the
predicted satellite number once $M_0 \ll M_1$. The tighter constaints are also 
particularly puzzling given that \citet{zehavi_etal11} adopt an arbitrary 5\% error on 
all galaxy number density measurements, an error that greatly exceeds the 
Poisson error, typically $< 1\%$, that we adopt. This additional error 
contribution also limits the value of comparing our $\chi^2$ values with those of 
\citet{zehavi_etal11}, as the latter values will be artificially lower. 
Therefore, it seems likely that the tighter constraints quoted by \citet{zehavi_etal11}
must be caused by an error.
